---
title: "DPLYR"
author: "Paulina Grunwald - www.R-HackRoom.com"
date: "December 21, 2016"
output: html_document
---

#####MANAGING AND MUNGING DATA IN R USING DPLYR AND TIDYR#####

###1. Using DPLYR in R

Dplyr is a new package which provides a set of tools for efficiently manipulating datasets in R. You can install and then load dplyr from the library.


```{r}
#Install dplyr package
install.packages("dplyr")
```

```{r setup, include=FALSE}
#Load dplyr from the library
library(dplyr)
```

In this example we will use nycflights13 dataset which can be installed from Cran repository.

```{r}
#Install nycflights13 dataset
install.packages("nycflights13", repos='http://cran.us.r-project.org')
```
```{r}
#Load nycflights13 from the library
library("nycflights13")
```

Now we will wrap our dataframe with tbls using special tbl dataframe - tbl_df . Tbl_df is function in dplyr and it used to prevents long table outputs when accidentally calling the data frame. Tbls also shows the size of the dataset () and will display amount of data that will fit in your console.

```{r}
flights <- tbl_df(flights)
#print flights dataframe
flights
```

Now we can check class of flights dataframe:
```{r}
class(flights)
```


Now we will display dimentions of the dataset:
```{r}
dim(flights)
```
As we can see the dataset has 19 collumns and 336776 rows.

Dplyr implements the following verbs useful for data manipulation:

* select(): focus on a subset of variables
* filter(): focus on a subset of rows
* mutate(): add new columns
* summarise(): reduce each group to a smaller number of summary statistics
* arrange(): re-order the rows


#Select all columns that end with time
```{r}
select(flights, ends_with("time"))
```

#Select columns that contain "_"
```{r}
select(flights, contains("_"))
```

#Select columns that contain delay 

```{r}
select(flights, contains("delay"))
```

#Instead writing code provided above we can use pipe operator %>%
dataset_name %>% select(starts_with("dep"))
Example:
```{r}
flights %>% select(ends_with("time"))
```

You can use keboard shorecut for %>%
Mac: Cmd+Shift+M
Win: Ctrl+Shift+M


###Selecting Rows from the dataframe###
dataset_name %>% filter(col1 > X)
dataset_name %>% filter(col1 %in% c("X", "Y"))

You can combain two expressions using comma:

dataset_name %>% filter(col1 > X, 
  col1 %in% c("X", "Y"))

##Logical tests in R:##
COMPARISON:
< less than
> greater than
== equal to
<=
>=
!= Not equal to
%in% Group membersit
is.na
!is.na
LOGIC:
& and
| or
! mpt
any any true
all all true

### DERIVE NEW VARIABLES FROM THE DATAFRAME

mutate() function
